---
title: "ChlA_Lab"
author: "BillyMcLean"
date: "2024-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(tidyverse)
```

```{r}
YojoaData <- read_csv("Chla_data_Yojoa_Enrichment.csv")

```

```{r}
YojoaData <- YojoaData %>% 
  mutate(Chl_a_ugL = as.numeric(Chl_a_ugL))


```

1.  What was the average chlorophyll content for the control bottles? Nitrogen addition? Phosphorus? Nitrogen and Phosphorus? What is the standard error of each? Results may be displayed in a table.\

```{r}

YojoaData_averages <- YojoaData %>% 
  group_by(Treatment) %>% 
  summarize(avg = mean(Chl_a_ugL))

YojoaData_sd <- YojoaData %>% 
  group_by(Treatment) %>% 
  summarize(sd = sd(Chl_a_ugL))

print(YojoaData_averages)
print(YojoaData_sd)

```

```{r}
ggplot(data = YojoaData)+
  geom_boxplot(mapping = aes(x = Treatment, y = Chl_a_ugL))+
  labs(x = "Treatment",
      y = "Chlorophyll A (ug/L)")
```

2.  Based on your calculations, what appears to be the primary source of limitation in Lake Yojoa in January: Nitrogen, Phosphorus or Co-limitation?

With an average of \~10.1, Phosphorus is the primary limiting nutrient. Nitrogen and Phosphorous could be considered co-limiting, with their combined average being \~11.5, but as Nitrogen is at 5.96, Phosphorous is the primary.

3.  What are other potential sources of limitation not accounted for in this experiment?

4.  Produce one figure or statistical test of your choice that you feel best supports your conclusions in question 2 (i.e. bar chat with error bars, box plots, ANOVA etc.)

```{r}
N <- filter(YojoaData, Treatment == "Nitrogen")
Ctrl <- filter(YojoaData, Treatment == "Control") 
P <- filter(YojoaData, Treatment == "Phosphorus")


t.test(N$Chl_a_ugL,P$Chl_a_ugL)
t.test(N$Chl_a_ugL,Ctrl$Chl_a_ugL)
t.test(P$Chl_a_ugL,Ctrl$Chl_a_ugL)



```

